\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid

### 2014-09-24 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Filled in header.
Major editting pre-Bill. Added in more numbers when discussing figures.

\ecjHeader{x}{x}{xxx-xxx}{200X}{TODO 45-character paper description}{B. W. Goldman and W. F. Punch}
\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}

@@

\begin{abstract}
%Should be about 200 words
The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
evolutionary optimization without requiring problem specific user specified parameters.
This is achieved by combining local search with advanced crossover to iteratively construct
and exploit multiple populations simultaneously, with each population akin to a generation
in traditional search.

When compared with five advanced search algorithms across dozens of problem sizes for
seven different benchmarks P3 reliably required at least a constant factor less evaluations to find the global optimum
than all tested competitors. P3 also in general maintains the highest fitness reached when comparing
algorithms based on different number of evaluations performed. Some evidence suggests this performance
also extends to considering total number seconds to find the global optimum.
We investigate the causes of this efficiency and provide some evidence that P3's ability to
add diversity only as needed may be contributing to its success.

@@

\section{Comparison Algorithm Parameter Tuning}
\label{sec-tuning}
While four of the six algorithms in our experiments do not require any user specified parameters,
hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
handicapped, we have extensively tuned each using the bisection method \citep{sastry:2001:bisection}
to determine the optimal population size for each size of each problem.
Extended by \cite{goldman:2012:ltga} this method iteratively
doubles the population size until some success criteria is met and then performs bisection
between the lowest successful and highest unsuccessful sizes. Thus the minimum population size
which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.

@@

Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
are treated as requiring more evaluations than any successful run. If the median run was not successful
no point is shown.

The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
sizes, but have difficulty scaling to larger problems.  Only on MAX-SAT are these optimizers competitive
at larger tested problem sizes. However this may just be an artifact of that the largest tested MAX-SAT
was an order of magnitude smaller than the largest size tested for most other problems. On the Deceptive
Trap and Deceptive Step Trap problem $1+(\lambda, \lambda)$ has some other interesting results. For
problems with 4 or less traps it performs significantly better than any other algorithm. This is likely
due to an artifact of the landscape as $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
flipping entire traps. For larger problems it becomes increasingly unlikely that mutation and crossover can
create enough changes to an individual trap to overcome deception without negatively changing others. This
behavior also causes high variance in success rate, as evident by the occasional successes on large problems.
This ability to successfully integrate modifications of more than a single bit generally appears to lead $1+(\lambda, \lambda)$
outperform the Hill Climber on all problems except Nearest Neighbor NK.

As expected, Parameter-less hBOA was generally efficient than hBOA using an optimal population size. The slopes
of the two lines are relatively similar across all problems, meeting with the theoretical expectation that
Parameter-less hBOA is no more than a logarithmic factor worse than hBOA.  The only problem which reversed
this trend was MAX-SAT, where Parameter-less hBOA actually outperformed hBOA. A potential explanation for
this upset is that hBOA is using the optimal population size for the problem class, not for each instance.
Therefore hBOA must use a population size sufficiently large to solve the hardest instances in the class.
Parameter-less hBOA on the other hand is able to solve easy instances using much smaller population sizes.
The net effect is therefore that by being more adaptable Parameter-less hBOA is able to achieve better median
results.

@@

LTGA, as the current state-of-the-art, was generally the fastest to find the global optimum of all comparison
algorithms. This is especially true on the four single instance problems where it often found the global
optimum in a lower order of complexity.  However on Nearest Neighbor NK and Ising Spin Glasses LTGA appears
to perform worse than hBOA.  This may be due to LTGA's inability to represent the overlapping nature of these problems.
On MAX-SAT LTGA only outperforms hBOA. Again, the likely cause to this is LTGA's requirement to have a
population size sufficiently large to solve the hardest problem in each class, reducing its effectiveness
on instances of median difficulty.

@@

Across all problem and problem sizes P3 generally finds the global optimum using less evaluations than
all of the comparison algorithms.  On Deceptive Trap P3 significantly outperforms both hill climbing
and LTGA by combining the best features of each, using hill climbing to optimize each trap and crossover
to perfectly mix traps. Deceptive Step Trap less amenable to local search with P3's quality effectively
tied with both LTGA and hBOA. The slight improvement here may be due to P3's ability to solve some runs
using smaller population sizes than LTGA and hBOA due to lucky initialization. Using local search on HIFF
likely saves P3 a generation over LTGA and hBOA as all pairwise subproblems will already be solved. Combined
with P3's ability to add diversity only as needed this likely explains why P3 is able to solve HIFF a constant
multiple faster than either population based technique. P3 likely solves its median run of Nearest Neighbor
NK and Ising Spin Glasses faster than both hBOA and LTGA again by using population sizes fit to each instance
and not fit to the entire problem class. On MAX-SAT, where the most effective solvers are Hill Climber and
$1+(\lambda, \lambda)$, P3 outperforms both likely due to its use of local search with a better ability to
overcome deception.

To more fully examine the hypothesis that its P3's ability to scale to differences in problem difficulty,
Figure~\ref{fig-evals-to-success-range} shows the upper and lower quartiles for P3 and LTGA. On each problem
LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
population size regardless of instance and progresses search generationally. In contrast P3 has a much
higher split, with some runs finishing very quickly. This added variability is generally on only the lower
end as P3's upper quartile is only worse than LTGA's upper quartile on Deceptive Step Trap and Nearest
Neighbor NK. A possible explanation for this is that as P3 has not pre-learned the amount of diversity required
to solve all instances does not take as risky behavior as LTGA.

@@

At almost every recording interval for every problem hBOA has the worst fitness. The only exceptions come when
the total number of evaluations performed is significantly less than the genome size hBOA reaches better fitness
than the Hill Climber, $1+(\lambda, \lambda)$, and P3. This early in search hBOA and LTGA are both still evaluating
individuals in their randomly generated initial populations, while the three techniques worse than hBOA have started
from a single solution and are attempting to optimize. Of the methods, hBOA uses by far the largest population sizes,
about 55 times larger than LTGA. As such it makes sense that hBOA would take a long time to being to reach high
quality solutions.

Parameter-less hBOA generally reports higher fitness than hBOA when using the same number of total evaluations. This makes
sense as Parameter-less hBOA's use of small populations allows it to begin to optimize long before hBOA has even evaluated
its entire initial population. However, this trend is reversed after a sufficient number of evaluations, as is apparent
on Deceptive Step Trap and Ising Spin Glasses. This makes sense as hBOA's population is set to quickly find the global
optimum.

Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This makes sense
as until P3 performs two restarts it is performing the same evaluations as the Hill Climber. Once P3 begins
performing crossover it immediately improves over the Hill Climber. As such P3 is likely better than a simple
hill climber regardless of how long each technique is run and irrespective of how high quality the solution
found has to be.

@@

For all problems, LTGA lags behind P3 in fitness improvement, but has a similar overall shape. On every problem except
Rastrigin, LTGA has a significant early delay in fitness improvement, followed by near instantaneous improvement, plateau,
and followed by a final period of improvement which results in the global optimum. This early period corresponds with
initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
This makes sense as LTGA's mixing strategy performs a sort of local search when the model has not yet detected gene epistasis.
This brings LTGA's fitness to almost the same level as the techniques explicitly performing local search, but is not quite
as effective at reaching local optima. Subsequent generations then make minor fitness improvements, but generally stagnate
while the model remains uninformed and high quality building blocks in the population are still relatively rare. Once the model
becomes accurate enough and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
enters a second period of rapid improvement.

@@

It is important to note that while P3 generally achieved the best fitness across all time points, hBOA and LTGA were tuned specifically
to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
global optimum, with the results given in Figure~\ref{fig-small-pop}. The two problems shown are representative of the behavior of using
a smaller population size on the other five problems. Reducing the population size resulted in LTGA's initial fitness improvement
occurring earlier, which makes sense as the first generation is completed much more quickly. On Deceptive Step Trap this means that
LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus. Across all tested problems using the smaller
population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap this prevents all 100
runs from reaching the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
which balances high quality intermediate fitness without trading away eventual optimality.

@@

Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding for search.
In order to calculate this value we recorded how many times search rebuilt the model during each run.
Under the minimal assumption that an evaluation costs $O(N)$ time and that model rebuilds cost $O(N^2)$,
Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically linear.
For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this appears to be the case. For both Trap problems
and HIFF there appears to be slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
bounded to no more than twice the algorithmic cost of an evaluation.

When applying a crossover subset, P3 tries random donors from the population until one is found with at least
one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
important note again is that each donation may range in size from a single bit up to $N-1$, with smaller
donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
in P3, but its unlikely to be very high.

@@

To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}. Due to the scale
of our experiments, diverse hardware potentially running multiple processes in parallel were used
in order to collect all runs. As such these timings should be considered very skeptically as a number
of sources may contribute to biased error in the results. Under those caveats, the results do suggest
that any increase in P3's algorithmic complexity is offset by its efficient use of evaluations. On
Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT it still appears to find the global
optimum faster than all comparison techniques. On HIFF LTGA is able to match P3's performance almost exactly,
and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
than LTGA but by no more than a constant factor.

@@

P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
that the relationship between the two is strongly problem dependent but relatively independent of problem size.
In general the trend appears to suggest that problems which can be more tightly represented by crossover tend
to require less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover about 60\%. The HIFF structure can
be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\%. Deceptive Step Trap spends the highest
percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
to mark all clusters as $useful$.

@@

In comparison with LTGA, the crossover success rates of P3 are lower but similar in problem ordering. This is almost certainly
a consequence of P3's use of local search. Counter intuitively the use of hill climbing on the initial population is likely reducing
P3's crossover success not because it reduces model quality or the donation pool, but because it is much more challenging to improve
locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
makes its aggregate crossover success incomparable to P3's.

@@

Another feature unique to P3 worth investigating is the shape and size of the population pyramid
constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
each level of the pyramid for the largest tested problem size for each problem. Each point is the
median size across 100 runs. If a run did not store any solutions at a level it is treated as 0,
and no point is drawn if the median run had 0 solutions stored at that level. This behavior is
representative pyramid shape across all problem sizes tested.

With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
in size as the level increases. This makes sense from an algorithmic standpoint as solutions are
added to the first level as long as they are unique local optima. For each subsequent level a
solution must be unique and a strict fitness improvement over its previously added version.
As a result it makes sense that it is generally less likely for a solution to be added to
higher levels than lower ones. This shape is also in many ways desirable from a population
genetics standpoint. When using a generational model \cite{lobo:2011:dynamicpop} found theoretical
evidence that the optimal population size reduces each generation with \cite{goldman:2011:dynamic-parameters}
providing supporting empirical data. In P3, levels of the pyramid are akin to generations as solutions
stored in each level have undergone the same number of variation and selection events,
with higher levels representing more of each. Therefore it is possible that P3's
pyramid shape is related to its ability to outperform competitors using an optimal
fixed population size. Subsequently smaller levels are able to focus search,
preserving only the diversity required by that stage of search. In comparison,
LTGA and hBOA set the population size at each generation equal to that required
to ensure sufficient initial diversity.

The number of solutions stored at the fourth level of Deceptive Step Trap is
significantly lower than that of the third or fifth levels, breaking the decreasing
trend of the other six problems. This behavior is rooted in the peculiar nature
of this landscape, and is also present in all other problem sizes tested.
After local search, all traps in all solutions have a total
number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
two bit plateaus, and as a result solutions in the second level are unlikely
to contain the lowest fitness local optima of 5 set bits. Similarly the third
level has very few traps set to the next worst local optima of 3 bits. As a result,
most solutions which reach the third level can only be improved by overcoming deception
and replacing the second best local optima of zero or one bits set to having all seven bits set.
At this point in search the population has not significantly increased the frequency of optimal
trap settings and with 8 ways to represent suboptimal local optima linkage is also likely to be
inaccurate. As a result our expectation is that when performing crossover with the third level
of the pyramid very few fitness improvements will actually be made, meaning
few solutions will be added to the fourth level. Those solutions that are added must, by definition,
contain a higher frequency of optimal trap settings to have achieved a fitness improvement. As
a result the crossover success rate and modeling quality at level four is likely to be much better
than at level three, increasing the likelihood of solutions being improved by crossover.

>>>>

% 45-character limit on header
\ecjHeader{x}{x}{xxx-xxx}{201X}{Parameter-less Population Pyramid}{B. W. Goldman and W. F. Punch}
\title{\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid}

@@

\begin{abstract}
%Should be about 200 words
The Parameter-less Population Pyramid (P3) is a recently introduced method for performing
evolutionary optimization without requiring problem specific parameters.
This is achieved by combining local search with advanced crossover to iteratively construct
and exploit multiple populations simultaneously, with each population akin to a generation
in traditional search.

When compared with five advanced search algorithms across dozens of problem sizes for
seven different benchmarks P3 reliably required at least a constant factor less evaluations to find the global optimum
than all tested competitors. When compared using fixed numbers of evaluations P3
generally finds the highest quality solutions. Some evidence also suggests this performance
extends to considering total number seconds to find the global optimum even when using efficient fitness functions.
We investigate the causes of this efficiency and provide some evidence that P3's ability to
add diversity only as needed may be contributing to its success.

@@

\section{Comparison Algorithm Parameter Tuning}
\label{sec-tuning}
While four of the six algorithms in our experiments do not require any user specified parameters,
hBOA and LTGA both use a population size parameter. To ensure these techniques are not unfairly
handicapped, we have extensively tuned each using the bisection method \citep{sastry:2001:bisection}
to determine the optimal population size for each size of each problem.
Extended by \cite{goldman:2012:ltga} this method iteratively
doubles the population size until some success criteria is met and then performs bisection
between the lowest successful and highest unsuccessful sizes. Thus the minimum population size
which meets the success criteria is found. \cite{goldman:2014:p3} proposed a success criteria
of performing $r$ successful runs in a row, such that the expected failure rate can be bounded
above by $\frac{3}{r+1}$~\citep{jovanovic:1997:ruleofthree}. As P3 and the other three algorithms
do not prematurely converge, we chose $r=100$ to similarly ensure hBOA and LTGA almost never do.
As bisection can make infinitely large population sizes, any run which had not found the global
optimum after 100 million evaluations or 128 computing hours was considered unsuccessful.

@@

Figure~\ref{fig-evals-to-success} shows the median number of evaluations required by each of the six
algorithms to find the global optimum for multiple sizes of each problem.  Each data point in
Figure~\ref{fig-evals-to-success} represents the median of 100 runs, where unsuccessful runs
are treated as requiring more evaluations than any successful run. If the median run was not successful
no point is shown. The maximum problem size used for each problem was set to be the largest size we could
feasibly determine an optimal population size for LTGA. For many larger problems results are not shown
for hBOA due to the extreme computational cost required to optimally set the population size.

The Random Restart Hill Climber and $1+(\lambda, \lambda)$ are both relatively effective on small problem
sizes, but have difficulty scaling to larger problems.  Only on MAX-SAT are these optimizers competitive
at larger tested problem sizes. However, this may just be an artifact as the largest tested MAX-SAT
was an order of magnitude smaller than the largest size tested for most other problems. On the Deceptive
Trap and Deceptive Step Trap problem $1+(\lambda, \lambda)$ has some other interesting results. For
problems with 4 or less traps it performs significantly better than any other algorithm. This is likely
because $1+(\lambda, \lambda)$ is able to overcome deception by probabilistically
flipping entire traps. For larger problems it becomes increasingly unlikely that mutation and crossover can
create enough changes to an individual trap to overcome deception without negatively changing others. This
behavior also causes high variance in success rate, as evident by the occasional successes on large problems.
This ability to successfully integrate modifications of more than a single bit generally appears to lead $1+(\lambda, \lambda)$
to outperform the Hill Climber on all problems except Nearest Neighbor NK.

As expected, Parameter-less hBOA was generally less efficient than hBOA using an optimal population size. The slopes
of the two lines are relatively similar across all problems, meeting with the theoretical expectation that
Parameter-less hBOA is no more than a logarithmic factor worse than hBOA.  The only problem which reversed
this trend was MAX-SAT, where Parameter-less hBOA actually outperformed hBOA. A potential explanation for
this upset is that hBOA is using the optimal population size for the problem class, not for each instance.
Therefore hBOA must use a population size sufficiently large to solve the hardest instances in the class.
Parameter-less hBOA on the other hand is able to solve easy instances using much smaller population sizes.
The net effect is therefore that by being more adaptable Parameter-less hBOA is able to achieve better median
results.

@@

LTGA, as the current state-of-the-art, was generally at least as fast to find the global optimum as hBOA.
However on Nearest Neighbor NK and Ising Spin Glasses LTGA appears
to perform worse than hBOA.  This may be due to LTGA's inability to represent the overlapping nature of these problems.
On MAX-SAT LTGA only outperforms hBOA and performs worse than the other 4 techniques.
Again, the likely cause to this is LTGA's requirement to have a
population size sufficiently large to solve the hardest problem in each class, reducing its effectiveness
on instances of median difficulty.

@@

Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
The largest of these used 49 bits, with P3 achieving the best results on all 92 larger
configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the 5, 4, and 1 smallest Deceptive
Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the 3 smallest Nearest Neighbor NK instances
and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
on 5 smallest Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.

On Deceptive Trap P3 significantly outperforms both hill climbing
and LTGA by combining the best features of each, using hill climbing to optimize each trap and crossover
to perfectly mix traps. Deceptive Step Trap is less amenable to local search with P3's quality effectively
tied with both LTGA and hBOA. The slight improvement here may be due to P3's ability to solve some runs
using smaller population sizes than LTGA and hBOA due to lucky initialization. Using local search on HIFF
likely saves P3 a generation over LTGA and hBOA as all pairwise subproblems will already be solved. Combined
with P3's ability to add diversity only as needed this likely explains why P3 is able to solve HIFF a constant
multiple faster than either population based technique. P3 likely solves its median run of Nearest Neighbor
NK and Ising Spin Glasses faster than both hBOA and LTGA again by using population sizes fit to each instance
and not fit to the entire problem class. On MAX-SAT, where the most effective solvers are Hill Climber and
$1+(\lambda, \lambda)$, P3 outperforms both likely due to its use of local search with a better ability to
overcome deception.

To more fully examine the hypothesis that some of P3's success is do to its ability to scale to differences in problem difficulty,
Figure~\ref{fig-evals-to-success-range} shows the upper and lower quartiles for P3 and LTGA. On each problem
LTGA has a much smaller difference between its best and worst runs. This makes sense as LTGA uses the same
population size regardless of instance and progresses search generationally. In contrast P3 has a much
higher split, with some runs finishing very quickly. This added variability is generally on only the lower
end as P3's upper quartile is only worse than LTGA's upper quartile on Deceptive Step Trap and Nearest
Neighbor NK. A possible explanation for this is that as P3 has not pre-learned the amount of diversity required
to solve all instances does not take as risky of behavior as LTGA.

@@

% 181 total
% 126 has P3 best
% 27 LTGA
% 20 hboa
% 50 lambdalambda
% hc 9
% phboa 18

At 63\% of recording intervals hBOA has the worst fitness of any solver. Many of the exception occur when
the total number of evaluations performed is significantly less than the genome size and hBOA reaches better fitness
than the Hill Climber, $1+(\lambda, \lambda)$, and P3. This early in search hBOA and LTGA are both still evaluating
individuals in their randomly generated initial populations, while the three techniques worse than hBOA have started
from a single solution and are attempting to optimize. Of the methods, hBOA uses by far the largest population sizes,
about 55 times larger than LTGA on average for problems shown. As such it makes sense that hBOA would take a long time to reach high
quality solutions.

Parameter-less hBOA reported equal or higher fitness than hBOA when using the same number of total evaluations in 68\% of intervals. This makes
sense as Parameter-less hBOA's use of small populations allows it to begin to optimize long before hBOA has even evaluated
its entire initial population. However, this trend is reversed after a sufficient number of evaluations, as is apparent
on Deceptive Step Trap and Ising Spin Glasses. This makes sense as hBOA's population is set to quickly find the global
optimum.

Early in optimization P3 and the Random Restart Hill Climber have effectively identical quality. This is because
P3 performs the same evaluations as the Hill Climber for the first two restarts. Once P3 begins
performing crossover it immediately improves over the Hill Climber. In 95\% of intervals P3 had a fitness at
least as high as Hill Climbing. As such P3 is likely better than a simple
hill climber regardless of how long each technique is run and irrespective of how high quality the solution
found has to be.

@@

For all problems, LTGA lags behind P3 in fitness improvement, but has a similar overall shape. On every problem except
Rastrigin, LTGA has five distinct periods: fitness plateau, near instantaneous improvement, fitness plateau,
and improvement to global optimum. The early period corresponds with
initialization of the population, with the first fitness gain achieved immediately upon completing the first generation.
This can be explained as LTGA's mixing strategy performs a sort of local search when the model has not yet detected gene epistasis.
This brings LTGA's fitness to almost the same level as the techniques explicitly performing local search, but it is not quite
as effective at reaching local optima. Subsequent generations then make minor fitness improvements, but generally stagnate
while the model remains uninformed and high quality building blocks in the population are still relatively rare. Once the model
becomes accurate enough and the probability of a crossover using high quality genetic material increases sufficiently, LTGA
enters a second period of rapid improvement.

@@

It is important to note that while P3 achieved the best fitness across 70\% of time points, hBOA and LTGA were tuned specifically
to find the global optimum and are not necessarily using the optimal population size for finding intermediate fitnesses. Consider
that for both reducing the population size would almost certainly improve their fitness for evaluation steps less than their population
size. To examine the effect of population size we tested LTGA using one tenth of the population size required for reliably finding the
global optimum, with the results given in Figure~\ref{fig-small-pop}. The two problems shown are representative of the behavior of using
a smaller population size on the other five problems. In aggregate, P3 still achieved a fitness at
least as high as LTGA using the smaller population size at 80\% of intervals.
Reducing the population size resulted in LTGA's initial fitness improvement
occurring earlier, as the first generation is completed much more quickly. On Deceptive Step Trap this means that
LTGA leaps ahead of P3 due to its ability to overcome the two bit fitness plateaus. Across all tested problems using the smaller
population size reduced the quality of LTGA's first fitness plateau, likely due to missing required diversity to reach higher quality.
The second period of improvement also comes earlier and reaches a lower quality fitness. On Deceptive Step Trap 0
runs reach the global optimum, down from 100 successful for the full population size. On Nearest Neighbor NK this drops the number
of successful runs from 98 to 68. As a result we conclude that while reducing the population size of LTGA may improve its quality
at early points during optimization, doing so reduces the robustness of the final solution found. This is in contrast to P3
which balances high quality intermediate fitness without trading away eventual optimality.

@@

Figure~\ref{fig-rebuilds} reports in an algorithmic sense how expensive model rebuilding is for search.
In order to calculate this value we recorded how many times search rebuilt the model during each run.
Under the minimal assumption that an evaluation costs $\Omega(N)$ time and that model rebuilds cost $O(N^2)$,
Figure~\ref{fig-rebuilds} shows the ratio cost of model rebuilding per evaluation. If the cost of model building
scaled linearly with evaluations, the relationship plotted for each problem should be asymptotically constant.
For Nearest Neighbor NK, Ising Spin Glasses, and Rastrigin this appears to be the case. For both Trap problems
and HIFF there appears to be slow growth in the ratio. The problem sizes used for MAX-SAT were not sufficient
to accurately gauge the asymptotic behavior. Together this suggests that while the cost of building the model
is almost linear per evaluation, it can grow slowly. However, even in the worst case (HIFF) this growth was
bounded to no more than twice the algorithmic cost of an evaluation.

When applying a crossover subset, P3 tries random donors from the population until one is found with at least
one bit different from the improving solution. In theory this can result in up to $O(\mu)$ operations.
Figure~\ref{fig-donations} examines the observed average number of donations per evaluation performed.
Ising Spin Glass, HIFF, and Rastrigin all achieve effectively constant behavior here, implying repeated
donation does not impact the asymptotic runtime of P3. Both Trap functions and Nearest Neighbor NK all
increase in number of donations as problem size increases, potentially increasing algorithmic costs. An
important note is that each donation may range in size from a single bit up to $N-1$, with smaller
donations far more likely to result in repeated attempts. As such this may cause some super-linear growth
in P3, but its unlikely to be very high.

@@

To assess wall clock performance we provide Figure~\ref{fig-seconds-to-success}. Due to the scale
of our experiments, diverse hardware potentially running multiple processes in parallel were used
in order to collect our results. As such these timings should be considered very skeptically as a number
of sources may contribute to biased error in the results. Under those caveats, the results do suggest
that any increase in P3's algorithmic complexity is offset by its efficient use of evaluations. On
Deceptive Trap, Nearest Neighbor NK, Ising Spin Glasses, and MAX-SAT it still appears to find the global
optimum faster than all comparison techniques. On HIFF LTGA is able to match P3's performance almost exactly,
and on Rastrigin P3 falls behind only on moderate sized problems. Deceptive Step Trap yields the worst performance,
with LTGA using optimal population sizes having a significantly lower growth curve than P3. This mirrors
the results from Figure~\ref{fig-evals-to-success} under the assumption that P3's computational time per evaluation is higher
than LTGA but by no more than a constant factor. In aggregate P3 found the global optimum the fastest on 72 of 130 configurations,
with LTGA, $1+(\lambda, \lambda)$, Hill Climber, hBOA outperforming P3 on 44, 18, 10, and 1 configurations, respectively.

@@

P3 performs optimization using a mixture of hill climbing and crossover. Exactly how much time the algorithm
dedicates to each is not clear from the algorithm description alone. As such, Figure~\ref{fig-cross} shows
that the relationship between the two is strongly problem dependent but relatively independent of problem size.
In general the trend appears to suggest that problems which can be more tightly represented by crossover tend
to require less crossover evaluations. Deceptive Trap and Rastrigin spent the least evaluations on crossover, using
15\% and 20\% respectively on the largest problem sizes. Both of these problems are perfectly separable and have relatively
few local optima in each subproblem. HIFF and Ising Spin Glasses use crossover for about 60\% of evaluations. The HIFF structure can
be perfectly learned by the linkage tree, but the higher order relationships require multiple stages of crossover to optimize.
Ising Spin Glasses have an overlapping structure, making them impossible to perfectly represent in a single linkage tree, but
that overlap is relatively minor in comparison with Nearest Neighbor NK, which uses 81\%. Deceptive Step Trap spends the highest
percentage of evaluations on crossover, 92\%. Unlike all of the other problems, the value taken by an individual bit in Deceptive
Step Trap has very little meaning. This is because any trap value is either a local optima or 1 bit flip away from being a local optima.
This has a combined result of meaning it takes very few local search evaluations to find a local optima but very many crossover operations
to proliferate the global optima. Epistatic relationships are also going to be very weak and even small population sizes are likely
to mark all clusters as $useful$.

@@

In comparison with LTGA, the crossover success rates of P3 are lower but similar in problem ordering. This is almost certainly
a consequence of P3's use of local search. Counterintuitively the use of hill climbing on the initial population is likely reducing
P3's crossover success, not because it reduces model quality or the donation pool, but because it is much more challenging to improve
locally optimal solutions than randomly generated ones. LTGA's crossover benefits from application to unoptimized solutions, which
makes its aggregate crossover success incomparable to P3's.

@@

Another feature unique to P3 worth investigating is the shape and size of the population pyramid
constructed for each problem. Figure~\ref{fig-level-size} shows the number of solutions stored at
each level of the pyramid for the largest tested problem size for each problem. Each point is the
median size across 100 runs. If a run did not store any solutions at a level it is treated as 0,
and no point is drawn if the median run had 0 solutions stored at that level. While pyramid size
is effected by problem size, the overall shape is not. As such the behavior shown in Figure~\ref{fig-level-size}
is representative of that for all other tested problem sizes.

With the exception of the dip in Deceptive Step Trap, all of the pyramids show a monotonic reduction
in size as the level increases. This makes sense from an algorithmic standpoint as solutions are
added to the first level as long as they are unique local optima. For each subsequent level a
solution must be unique and a strict fitness improvement over its previously added version.
As a result it makes sense that it is generally less likely for a solution to be added to
higher levels than lower ones. This shape is also in many ways desirable from a population
genetics standpoint. When using a generational model \cite{lobo:2011:dynamicpop} found theoretical
evidence that the optimal population size reduces each generation and \cite{goldman:2011:dynamic-parameters}
made similar empirically based conclusions. In P3, levels of the pyramid are akin to generations as solutions
stored in each level have undergone the same number of variation and selection events,
with higher levels representing more of both. Therefore it is possible that P3's
pyramid shape is related to its ability to outperform competitors using an optimal
fixed population size. Subsequently smaller levels are able to focus search,
preserving only the diversity required by that stage of search. In comparison,
LTGA and hBOA set the population size at each generation equal to that required
to ensure sufficient initial diversity.

The number of solutions stored at the fourth level of Deceptive Step Trap is
significantly lower than that of the third or fifth levels, breaking the decreasing
trend of the other six problems. This behavior is rooted in the peculiar nature
of this landscape, and is also present in all other problem sizes tested.
After local search, all traps in all solutions have a total
number of 1 bits equal to \{0, 1, 3, 5, 7\} as these correspond to the local
optima when using $k=7$ and $s=2$. Crossover is very likely to overcome the
two bit plateaus, and as a result solutions in the second level are unlikely
to contain the lowest fitness local optima of 5 set bits. Similarly the third
level has very few traps set to the next worst local optima of 3 bits. As a result,
most solutions which reach the third level can only be improved by overcoming deception
and replacing the second best local optima of zero or one bits set to having all seven bits set.
At this point in search the population has not significantly increased the frequency of optimal
trap settings and with 8 ways to represent suboptimal local optima linkage is also likely to be
inaccurate. As a result our expectation is that when performing crossover with the third level
of the pyramid very few fitness improvements will actually be made, meaning
few solutions will be added to the fourth level. Those solutions that are added must, by definition,
contain a higher frequency of optimal trap settings to have achieved a fitness improvement. As
a result the crossover success rate and modeling quality at level four is likely much better
than at level three, increasing the likelihood of solutions being improved by crossover.

When examining the crossover success rate shown in Figure~\ref{fig-level-success} these suppositions are
supported. The success rate for Deceptive Step Trap decreases for the first three levels, with the third level
successful only 0.0004\% of the time. This quickly rebounds at higher levels. On all other problems, the
crossover success rate follows a trajectory closer to expectation. Low levels are more successful than medium
levels as it is easier to make improvements to less optimized solutions. This trend reverses somewhat at
higher levels as solution filtering allows for highly accurate models and high frequencies of high quality
building blocks. The highest level on most problems has a very low crossover success rate, likely because
so few solutions other than the global optimum every cross with that level.
